The Creation of a Programming Language

### 2022-10-25 paper.tex
continue writing paper

\section{Implementation}
\subsection{Used Software}
At the heart of making any project in the C programming language
stands the C compiler. This project uses
\emph{gcc}, the GNU Compiler Collection, which, as the name suggests, also 
supports more languages than solely C. Substantiating the development process is 
\emph{GNU Make}, a general-purpose build system. It handles tasks like building
the final executable program step-by-step from the project source code and directory clean-up.
The C ecosystem is very easy to use on and fundamentally supported by 
\emph{GNU+Linux}, the operating system used to program on.
The code editor used was \emph{Microsoft Visual Studio Code}, providing great
help with built-in git integration and auto-completion features.
\emph{Git} is the project management and tracking software used to back up and share
the project files on \emph{GitHub.com}, a Microsoft hosted git repository and
web front-end to store and collaborate on source code.

@@

\subsubsection{The Bytecode Compiler}
The bytecode compiler is an internal part of the interpreter, located in the file
\hrc{src/interpreter/bytecode.c}. Its job is to create a list
of byte-sized instructions and data from the syntax-tree the parser created.
Having such an intermediary bytecode format, instead of directly interpreting
the parse-tree can be beneficial. This is the case when the same piece of code
has to be run several times. Instead of generating a time and memory overhead 
having to walk along tree-nodes recursively, only the list of instructions in
the bytecode can be traversed, which is much faster.

Not to be confused with machine-native bytecode, the bytecode compiler in \name{}
generates byte-sequences in an internal format, that are then interpreted. A
JIT compiler would instead compile the AST down to machine-native bytecode, which
could run directly on the system's hardware and would not have to be interpreted.

To generate such a sequence of bytes, \name{} recursively reads through the
binary tree created by the parser and pushes found instructions onto a stack,
which is in this case used similarly to a variable-sized list.
The algorithm used is inspired by Edsger Dijkstra's 'Shunting yard 
algorithm', traversing the tree post-order to create
a prefix notation. 
[\cte{shunting_yard}]

@@

In \hrc{src/interpreter/bytecode.c} we can see that first the left operand and then
the right operand are added onto the stack of bytecode. Therefore, when
popping from the stack and evaluating instructions, the operands will have to
be read in reverse.
Here is an example:
\begin{enumerate}
    \item Assume the following mathematical expression in infix notation:
        \expr{\emph{'1 - 2 * 4 / (5 + 6)'}}
    \item In common prefix notation, also called Polish notation, the
        operators would now precede the operands in this manner:
        \expr{\emph{'- 1 (/ (* 2 4) (+ 5 6))'}}
    \item But in \name{} the operands are pushed onto the stack is reverse order.
        This would look like the following:
        \expr{\emph{'- (/ (+ 6 5) (* 4 2)) 1'}}
    \item The brackets are not needed, because for every operator, the amount
        of operands is known. This is the case as e.g. the subtraction operator
        and the negative operator, which are both a minus sign, are distinguished
        beforehand. This means the above statement would look like this:
        \expr{\emph{'- / + 6 5 * 4 2 1'}}
        This is an ideal abstraction of how data 
        is represented internally in \name{}.
\end{enumerate}
[\cte{polish_notation}]

\subsubsection{The Runtime}
This is the last stage of interpreting program code as implemented in \name{}.
The runtime steps through the intermediary bytecode created by the bytecode
compiler and evaluates expressions by plugging in the provided values.
All available expressions are implemented in 
\hrc{src/interpreter/processing/implementations.c}.

The evaluation of expressions also works in a recursive manner. Here is an 
example of \name{}'s bytecode notation and how it would be evaluated at runtime:
\begin{itemize}
    \item Assume the input in \name{}'s bytecode to resemble the following expression: 
        \expr{\emph{'- 6 * 3 4'}}
        In infix notation, this would be equivalent to 
        \expr{\emph{'(4 * 3) - 6'}.}
    \item The runtime will read through the bytecode left-to-right. First it will
        notice the subtraction-operator and accordingly assume that two
        operands must follow.
    \item The processing function will call itself
        and notice the 6, a number, and therefore a value. The function simply returns the six
        to the subtraction expression. This six will be the right-hand operator
        of the subtraction.
    \item The subtraction will then call the processing function again to find
        the left operand. As the processing function reads a multiplication operator,
        it now recursively reads through that expression's right and then left
        operator, which would be 3 and for respectively and compute and return
        the result of the multiplication, which is 12.
    \item Now the subtraction knows its right- and left-hand operators, 6 and 12,
        and again, calculate and return the result, which is 6.
\end{itemize}

That is it. With all expressions that are currently implemented, entire programs
can be interpreted. Examples of such programs can be found in the \emph{examples} directory
of the source code.

@@

% TODO: add examples in /examples/
\name{} is built upon the implementation of a recursive Pratt
parser and a basic tokenizer without which the entire \name{} implementation
would be unusable.

@@

\subsection{Name Shadowing}
Here is another example, which in contrast to the example above, focuses on a
security concept implemented in \name{}.
\code{examples/exec.nts}{1}{999}
The code begins by creating the variable \emph{var}, which holds a string, and
\emph{f}, which calls the function \emph{std.exec} with var as argument.
Line number six holds a comment, indicated by the two slashes at the start of the line.
Comments are ignored by the \name{} interpreter and their purpose is to contain notes
and explanations inside the source code itself. Another type of commends are block-comments
which span across multiple lines, such as on lines 9 to 12.
On line seven, the \emph{std.exec} function is called for the first time. It runs the
command it is given as argument in the user's shell, which is most commonly \emph{cmd} on
Microsoft Windows and \emph{bash} on most Linux distributions. In this example, \emph{std.exec}
launches the Firefox browser, but it is capable of all kinds of things that can otherwise be
done in the shell, such as delete files or shut down the computer.

@@

\subsection{Limitations}\label{Limitations}
\name{} can be called a programming language without a doubt. But it is not perfect in any
way. There are limitations that make it practically unusable for actual
professional projects. In this section, a few shortcomings of \name{} are listed
and reasoned over.

@@

Other programming languages, notably object-oriented languages such as Java,
use Garbage-Collection, which automatically tracks unused allocated memory and
frees it up automaticadlly. While this is also the case with \name{}, there are
many internal flaws that have nothing to do with variables used in a \name{}
program, but rather internally.

A major reason for the memory leaks, but not an excuse, is the nature of the
C programming language the \name{} interpreter was written in. The C language and
compiler do not keep track of states in which certain blocks of memory are
forgotten about and warn the developer, instead, memory leaks happen.
More modern C alternatives such as Rust have a ownership-system in place,
in which any allocated memory has a function as an owner, and when the function
is done processing, the memory is automatically freed. Such systems are a
helpful tool for developers to create better programs with low mental overhead,
and are what programming language design should be all about.

\subsubsection{Single-Threaded}
The \name{} interpreter does not implement any multi-threaded, meaning parallel
running, programs as most other languages do. Instead, any \name{} program can
only be interpreted and run sequentially. Since most modern computer processors
are equipped with a handful of processing cores and threads, not being able to
make use of those is a major downside and in itsself a reason not to use this
language.

\subsubsection{Lack of a Development Environment}
This is certainly the greatest problem of \name{}, aside the countless smaller
bothering aspects.

Most modern programming languages do not only consist
of a language specification and a compiler or interpreter, but of many more
tools surrounding them.
Such tools can include a package-manager, with which people around the world
can share code. Examples would be npm or cargo, 
the 'Node Package Manager' for JavaScript applications and and Rust's package 
manager respectively.
[\cte{npm}, \cte{cargo}]

@@

\subsection{Conclution}
Creating an interpreter was a challenge. From the beginning on, the project
had to be planned and commented strictly, not to get lost in the complex web
of source code later on. The final product is a success, a working interpreter
hosting a custom programming language that can run human-readable programs.
Although there are several limitations and shortcomings as stated in section \ref{Limitations}
of this paper, the details of the implementation brought about unexpected 
problems and solutions to profit from. The goal of this paper, to learn about 
the inner workings of an interpreter, has therefore been achieved.

>>>>

\section{Implementation}
\subsection{Used Software}
At the heart of making any project in the C programming language
stands the C compiler. This project uses
\emph{gcc}, the GNU Compiler Collection, which, as the name suggests, also 
supports more languages than solely C. Substantiating the development process is 
\emph{GNU Make}, a general-purpose build system. It handles tasks like building
the final executable program step-by-step from the project source code and directory clean-up.
The C ecosystem is very easy to use on and fundamentally supported by 
\emph{GNU+Linux}, the operating system used to program on.
The code editor used was \emph{Microsoft Visual Studio Code}, providing great
help with built-in git integration and auto-completion features.
\emph{Git} is the project management and tracking software used to back up and share
the project files on \emph{GitHub.com}, a Microsoft hosted git repository and
web front-end to store and collaborate on source code.
% TODO: add gdb!

@@

\subsubsection{The Bytecode Compiler}
The bytecode compiler is an internal part of the interpreter, located in the file
\hrc{src/interpreter/bytecode.c}. Its job is to create a list
of byte-sized instructions and data from the syntax-tree the parser created.
Using such an intermediary bytecode format instead of directly interpreting
the parse-tree can be beneficial. This is the case when the same piece of code
has to be run several times. Instead of generating a time and memory overhead 
having to walk along tree-nodes recursively, only the list of instructions in
the bytecode can be traversed, which takes less time overall.

Not to be confused with machine-native bytecode, the bytecode compiler in \name{}
generates byte-sequences in an internal format that are then interpreted. A
JIT compiler would instead compile the AST down to machine-native bytecode, which
could run directly on the system's hardware and would not have to be interpreted.

To generate the bytecode, \name{} recursively reads through the
binary tree created by the parser and pushes found instructions onto a stack.
The algorithm used is inspired by Edsger Dijkstra's 'Shunting yard 
algorithm', traversing the tree post-order to create
a prefix notation. 
[\cte{shunting_yard}]

@@

In \hrc{src/interpreter/bytecode.c} we can see that first the left operand and then
the right operand are added onto the stack of bytecode. Therefore, when
popping from the stack and evaluating instructions, the operands will have to
be read in reverse.
Here is an example:
\begin{enumerate}
    \item Assume the following mathematical expression in infix notation:
        \expr{\emph{1 - 2 * 4 / (5 + 6)}}
    \item In common prefix notation, also called Polish notation, the
        operators would now precede the operands in this manner:
        \expr{\emph{- 1 (/ (* 2 4) (+ 5 6))}}
    \item But in \name{} the operands are pushed onto the stack is reverse order.
        This would look like the following:
        \expr{\emph{- (/ (+ 6 5) (* 4 2)) 1}}
    \item The brackets are not needed, because for every operator, the amount
        of operands is known. This is the case as e.g. the subtraction operator
        and the negative operator, which are both a minus sign, are distinguished
        beforehand. The above statement would therefore look like this:
        \expr{\emph{- / + 6 5 * 4 2 1}}
        This is how expressions are represented in the bytecode of \name{}.
\end{enumerate}
[\cte{polish_notation}]

\subsubsection{The Runtime}
This is the last stage of interpreting program code as implemented in \name{}.
The runtime steps through the intermediary bytecode created by the bytecode
compiler and evaluates expressions by plugging in the provided values.
All available operators are implemented in 
\hrc{src/interpreter/processing/implementations.c}.

The evaluation of expressions also works in a recursive manner. Here is an 
example of \name{}'s bytecode notation and how it is evaluated at runtime:
\begin{itemize}
    \item Assume the input in \name{}'s bytecode to resemble the following expression: 
        \expr{\emph{- 6 * 3 4}}
        In common infix notation, this would be equivalent to 
        \expr{\emph{(4 * 3) - 6}.}
        The question arises, how can an algorithm compute the result of this
        calculation? The answer is recursion.
    \item The runtime will read through the bytecode left-to-right. First it will
        notice the subtraction-operator and accordingly assume that two
        operands must follow.
    \item The function recursively calls itself in order to evaluate the
        right operand and notice the 6, a value. The function simply returns the six
        to the subtraction expression.
    \item Again, the function calls itself to find the next operand.
        As it now reaches a multiplication operator,
        it recursively reads through that sub-expression's right and then left
        operator, which are 3 and 4 respectively.
        The result is computed and returned.
    \item Now the subtraction knows its right- and left-hand operators, 6 and 12.
        The subtraction is performed and the result, 6, is returned.
\end{itemize}

That is it. With all the operators and standard functions that have been implemented, entire programs
can be interpreted. Examples of such programs can be found in the \emph{examples} directory
of the source code.

@@

% TODO: add examples in /examples/
\name{} is built upon the implementation of a recursive Pratt
parser and a basic tokenizer. It uses a hash table to map variable names to their
memory addresses. There are no keywords, only a set of non-alphanumeric characters
representing all possible operations. Extending upon that, there are some
standard functions.

@@

\subsection{Name Shadowing}
Here is another example, which in contrast to the example above, focuses on an
original security concept implemented in \name{}.
\code{examples/exec.nts}{1}{999}
The code begins by creating the variable \emph{var}, which holds a string and
\emph{f}, which calls the function \emph{std.exec} with var as argument.
Line number six holds a comment, indicated by the two slashes at the start of the line.
Comments are ignored by the \name{} interpreter and their purpose is to contain notes
and explanations inside the source code itself. Another type of commends are block-comments
which span across multiple lines, such as on lines 9 to 12.
On line seven, the \emph{std.exec} function is called for the first time. It runs the
command it is given as argument in the user's shell, which is most commonly \emph{cmd} on
Microsoft Windows and \emph{bash} on most Linux distributions. In this example, \emph{std.exec}
launches the Firefox browser, but it is capable of all kinds of things that can otherwise be
done in the shell, such as delete files or shut down the computer.

@@

\subsection{Limitations}\label{Limitations}
Unfortunately, there are limitations that make \name{} practically unusable for actual
professional projects. In this section, a few shortcomings of \name{} are listed
and discussed.

@@

Other programming languages, notably object-oriented languages such as Java,
use Garbage-Collection, which automatically tracks unused allocated memory and
frees it up automatically. While this is also the case with \name{}, there are
many internal flaws that have nothing to do with variables used in a \name{}
program, but rather internally.

One reason memory leaks is the nature of the
C programming language, in which the \name{} interpreter was written in. The C language and
compiler do not keep track of all possible states in which certain blocks of memory are
forgotten about and do not warn the developer. Instead, memory leaks happen.
More modern C alternatives such as Rust have an ownership-system in place,
in which any allocated memory is owned by a function. When the function
is done processing, the memory is automatically freed. Such a system
is a helpful tool for developers to create better programs without having to get
lost in the details.
Inventing new ways to make the job of developers easier is what programming
language design is all about.

\subsubsection{Single-Threaded}
The \name{} interpreter does not implement any multi-threaded, i.e. in parallel
running programs as most other languages do. Instead, any \name{} program can
only run sequentially. Since most modern computer processors
are equipped with a handful of processing cores and threads, not being able to
make use of this modern time's advantage is a major downside.

\subsubsection{Lack of a Development Environment}
This is certainly the greatest problem of \name{}.

Most modern programming languages not only consist
of a language specification and a compiler or interpreter, but of many more
tools that surround themn.
Such tools can include a package-manager, with which people around the world
can share code. Examples would be npm or cargo, 
the 'Node Package Manager' for JavaScript applications and and Rust's package 
manager respectively.
Another example of a helpful tool is a debugger, helping developers
find the cause of errors, so-called bugs, in their programs.
This includes the tool gdb, the GNU Project Debugger, that was used in this
project to find errors in the \name{} implementation.
[\cte{npm}, \cte{cargo}]

@@

\subsection{Conclusion}
Creating an interpreter was a challenge. From the beginning on, the project
had to be planned and commented strictly, not to get lost in the complex web
of source code later on. The final product is a success, a working interpreter
hosting a custom programming language that can run human-readable programs.
Most parts of the source code are annotated with comments, giving interested
reader direct explanations of what is going on.
Although there are several limitations and shortcomings as stated in section \ref{Limitations}
of this paper, the details of the implementation brought about unexpected 
problems and solutions. The goal of this paper to learn about 
the inner workings of an interpreter has been achieved, as well as the goal
to get to know the C language.
